Left-linear rewriting in adhesive categories

### 2024-04-22 concur/fusione.tex
typos

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm,
% i.e. where the behaviour is dictated by a set of rules that have to be instantiated with the system at hand,
concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

@@

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, %to the best of our knowledge 
the phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{behr2021concurrency,BehrHK23}.

@@

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In we \S\ref{sec:equi} present the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class.
Finally, in \S\ref{conclusions} we outline possible directions for future work.
%
The appendices
% collect additional data,
recall basic results on adhesive categories (Appendix~\ref{app:ade}),
present \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds (Appendix~\ref{app:fill}),
and give full proofs of the results
(Appendix~\ref{omitted}).

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This first section recalls some definitions and the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so that
``$X\in \X$'' means that $X$ belongs to the class of objects of $\X$.
We will use $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$, assuming that $[0,n]=\emptyset$ whenever
$n<0$.

@@

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property of $\mathcal{M}$-adhesive categories is the \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. % We
% will recall it and examine some of its consequences.
In order to define it we need to introduce some terminology.
% and facts regarding subclasses of
% $\mor(\X)$.
%
%\begin{definition}
  Let  $\X$ be a category. A subclass $\mathcal{A}$ of
  $\mor(\X)$ is called\todo{il diagramma confonde la composizione}
  \begin{itemize}
    \parbox{11cm}{\item
      \emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
    \item \emph{closed under composition} if $g, f\in \mathcal{A}$ implies $g\circ f\in \mathcal{A}$ whenever $g$ and $f$ are composable.{\tiny }}\hfill
    \parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
    \parbox{11cm}{}\hfill
  \end{itemize}
%\end{definition}

@@

  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider\todo{diagramS?} the diagrams\\
    %
    \parbox{10cm}{
      aside.
      % 
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
    is an \emph{$\mathcal{A}$-Van Kampen square} if
      \begin{enumerate}
      \item it is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
      if and only if the front and right faces are pullbacks.
      \end{enumerate}}
    \parbox{2cm}{$\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D }$ }\\
    %    
    Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

@@

\begin{definition}[$\mathcal{M}$-adhesive category]
  Let $\X$ be a category and $\mathcal{M}$ a subclass of
  $\mon(\X)$
  % \begin{enumerate}
  % \item $\mathcal{M}$
  including  all isomorphisms, closed under composition,
  % \item $\mathcal{M}$
  and stable under pullbacks and pushouts.
  % \end{enumerate} 
  The category
  % 
  $\X$ is said to be \emph{$\mathcal{M}$-adhesive} if
  \begin{enumerate}
  \item it has \emph{$\mathcal{M}$-pullbacks}, i.e. pullbacks along arrows of $\mathcal{M}$;
  \item it has \emph{$\mathcal{M}$-pushouts}, i.e. pushouts along arrows of $\mathcal{M}$;
  \item  $\mathcal{M}$-pushouts are $\mathcal{M}$-Van Kampen squares.
  \end{enumerate}

@@

\begin{theorem}
  \label{thm:slice-functors}
  Let $\X$ be an $\mathcal{M}$-adhesive category. Given an object $X$
  the category $\X/X$ is $\mathcal{M}/X$-adhesive with
  $\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in
  \mathcal{M}$. Similarly, $X/\X$ is $X/\mathcal{M}$-adhesive with
  $X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in \mathcal{M}$.

@@

\begin{example}
  \label{ex:adhesive}
  $\cat{Set}$ is adhesive, and, more generally, every topos is
  adhesive~\cite{lack2006toposes}. By the closure properties above, every presheaf $[\cat{X},\cat{Set}]$ is adhesive, thus the category
  $\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is adhesive
  where $E \rightrightarrows {V}$ is the two objects category with two
  morphisms $s,t \colon{E} \to {V}$. Similarly, various
  categories of hypergraphs can be shown to be adhesive, such as term
  graphs and hierarchical graphs~\cite{CastelnovoGM24}. Note that the category $\cat{sGraphs}$ of simple graphs, i.e.
  graphs without parallel edges, is
  $\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not
  quasiadhesive.
\end{example}

@@

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair  of
  arrows $\rho = (l: K \to L, r: K \to r)$ such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

@@

  \vspace{.1cm}
  \noindent
  \parbox{10cm}{\hspace{15pt}Given two objects $G$ and $H$ and a rule
    $\rho=(l,r)$ in $\R$, a \emph{direct derivation $\mathscr{D}$ from
      $G$ to $H$ applying rule $\rho$} is a diagram as the one
    aside, in which both squares are pushouts. The arrow $m$ is called
    the \emph{match} of the derivation, while $h$ is its
    \emph{back-match}.  We denote a direct derivation $\dder{D}$
    between $G$ and $H$ as $\dder{D}\colon G\Mapsto H$. }
  \parbox{3cm}{$\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}&
      H}$}
\end{definition}

@@

%It can be seen that 
$\mathcal{M}$-adhesivity of $\X$ guarantees the
uniqueness of the result of applying a rule to an object, i.e. two
direct derivations using the same match are abstraction equivalent
(see Proposition~\ref{prop:unique}).

@@

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  We say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  \emph{sequentially independent}
  if there is pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes, referred to as an  \emph{independence
    pair}
  \[\xymatrix@R=16pt@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}
  \]
\end{definition}

@@

\begin{definition}[Switch]
  \label{def:switch}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation made
  by two sequentially independent derivations $\dder{D}_0$,
  $\dder{D}_1$ with independence pair $(i_0, i_1)$, as in the diagram
  on the left below. A \emph{switch} of $\der{D}$ along $(i_0,i_1)$ is
  a derivation $\der{E}=\{\dder{E}_i\}_{i=0}^1$, between the same objects and using the same rules
  in reverse order, as in the diagram on the right below
  \[
    \def\objectstyle{\scriptstyle}\def\labelstyle{\scriptstyle}
    \xymatrix@C=1.1mm@R=8mm{
      {L_0} \ar[d]_{m_0}
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]_{k_0}
      & & {R_0} \ar[drr]|(.35)\hole_(.6){h_0}  \ar@/^.22cm/[drrrrrr]|(.3)\hole^(.75){i_0}
      & & & & 
      % 
      {L_1}\ar[dll]|(.35)\hole^(.6){m_1} \ar@/_.22cm/_(.75){i_1}[dllllll]
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]^{k_1}
      & & {R_1} \ar[d]^{h_1} \\
      % 
      % 
      {G_0}
      & & {D_0} \ar@{>->}[ll]^{f_0} \ar[rrrr]_{g_0}
      & & & & {G_1} & &
      & &  {D_1} \ar@{>->}[llll]^{f_1} \ar[rr]_{g_1}
      & & {G_2}
    }
    % 
    \xymatrix@C=1.1mm@R=8mm{
      {L_1} \ar[d]_{m_0'}
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]_{k_0'}
      & & {R_1} \ar[drr]|(.35)\hole_(.6){h_0'}  \ar@/^.22cm/@{.>}^(.75){i_0'}|(.3)\hole[drrrrrr]
      & & & & 
      % 
      {L_0}\ar[dll]|(.35)\hole^(.6){m_1'} \ar@/_.22cm/@{.>}_(.75){i_1'}[dllllll] 
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_1'}
      & & {R_0} \ar[d]^{h_1'} \\
      % 
      % 
      {G_0}
      & & {D_0'} \ar@{>->}[ll]^{f_0'} \ar[rrrr]_{g_0'}
      & & & & {G'_1} & &
      & &  {D_1'} \ar@{>->}[llll]^{f_1'} \ar[rr]_{g_1'}
      & & {G_2}  }
    %
  \]
  such that there is an independence pair $(i_0', i_1')$ between
  $\dder{E}_0$ and $\dder{E}_1$ and 
  \begin{center}   
    $m_0=f_0' \circ i_1'$
    \qquad $h_1=g_1' \circ i_0'$
    \qquad $m_0'= f_0 \circ i_1$
    \qquad $h_1'= g_{1}\circ i_0$.
  \end{center}

@@

As we mentioned, for linear rewriting systems
%when viewing  derivations as concurrent computations, 
it is canonical to identify
derivations that are equal ``up to switching'', i.e. that differs
only for the order of independent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

@@

    Let us denote by $\nu_{k,h}$ the composition
  $\nu_h \circ \nu_{h-1} \circ \ldots \nu_k$. We say that the
  switching sequence consists of inversions if for all
  $k \in \interval[0]{m}$ the transposition $\nu_k$ is an inversion
  for $\nu_{1,m}$.

@@

Even if we work in strong enforcing rewriting systems where sequential
independence ensures switchability, when dealing with left-linear
rules there is a further -- and possibly more serious -- issue, i.e.
the fact that there can be more than one independence pair between
the same derivations. This hinders the very idea of using
sequential independence as a basis of a theory of concurrency for
rewriting systems, since exchanges performed using different
independence pairs may lead to derivations that are not abstraction
equivalent, and thus one would equate computations that should
definitively be taken apart, as shown in the example below.

\begin{example}
  Consider the derivation $\der{E}$ from Example~\ref{ex:seq-ind}.
  %
  The last two steps are sequential independent, but one easily sees
  that there are two distinct independence pairs, as the left-hand side
  of $\rho_1$ can be mapped either to node $1$ or to node
  $2$ in $G_2'$. Correspondingly, there
  are two switches of $\der{E}$: one is the derivation $\der{D}$ in
  Fig.~\ref{fi:derD} we started from, the other is the derivation
  $\der{D}'$ in Fig.~\ref{fi:derD1}.

@@

\begin{definition}[Node-merging graphical rewriting systems]
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph
  and $\G$ the free category over $\mathcal{G}$. The
  category $\gph{G}$ of \emph{$\mathcal{G}$-graphs} is the category
  $\Set^{\G}$. A
  \emph{node-merging graphical rewriting system}\todo{altro nome? source/leaf preserving} is a left-linear
  rewriting system $(\gph{G}, \R)$ such that  the
  following hold for each rule
  $(l\colon K\to L, r\colon K\to R)$ in $\R$
  \begin{enumerate}
  \item for every $v\in V$ and $x\in L(v)$, there exists a vertex $w$
    with depth $0$ and an arrow $p\colon w\to v$ (i.e.~a path in
    $\mathcal{G}$) such that $x$ belongs to the image of
    $L(p)\colon L(w)\to L(v)$;
  \item $r\colon K\to R$ is mono on vertices of depth $0$, i.e. for every
    $v\in V$ such that $\dph(v)=0$, the component $r_v:K(v)\to R(v)$
    is mono.
  \end{enumerate}
\end{definition}

>>>>

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm,
% i.e.~where the behaviour is dictated by a set of rules that have to be instantiated with the system at hand,
concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

@@

Technically, to acquire the expressiveness needed for capturing
fusions, i.e.~the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, %to the best of our knowledge 
the phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{behr2021concurrency,BehrHK23}.

@@

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class.
Finally, in \S\ref{conclusions} we outline possible directions for future work.
%
The appendices
% collect additional data,
recall basic results on adhesive categories (Appendix~\ref{app:ade}),
present \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds (Appendix~\ref{app:fill}),
and give full proofs of the results
(Appendix~\ref{omitted}).

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This first section recalls some definitions and the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so that
``$X\in \X$'' means that $X$ is an object of $\X$.
We will use $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$; in particular, $[0,n]=\emptyset$ whenever
$n<0$.

@@

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property of $\mathcal{M}$-adhesive categories is the \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. % We
% will recall it and examine some of its consequences.
In order to define it we need to introduce some terminology.
% and facts regarding subclasses of
% $\mor(\X)$.
%
%\begin{definition}
  Let  $\X$ be a category. A subclass $\mathcal{A}$ of
  $\mor(\X)$ is called %\todo{il diagramma confonde la composizione}
  \begin{itemize}
    \parbox{11cm}{\item
      \emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
    \item \emph{closed under composition} if $h, k\in \mathcal{A}$ implies $h\circ k\in \mathcal{A}$ whenever $h$ and $k$ are composable.{\tiny }}\hfill
    \parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
    \parbox{11cm}{}\hfill
  \end{itemize}
%\end{definition}

@@

  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider
    %\todo{diagramS?} 
    the diagram\\
    %
    \parbox{10cm}{
      aside.
      % 
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
    is an \emph{$\mathcal{A}$-Van Kampen square} if
      \begin{enumerate}
      \item it is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
      if and only if the front and right faces are pullbacks.
      \end{enumerate}}
    \parbox{2cm}{$\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D }$ }\\
    %    
    Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

@@

\begin{definition}[$\mathcal{M}$-adhesive category]
  Let $\X$ be a category and $\mathcal{M}$ a subclass of
  $\mon(\X)$
  % \begin{enumerate}
  % \item $\mathcal{M}$
  including  all isomorphisms, closed under composition,
  % \item $\mathcal{M}$
  and stable under pullbacks and pushouts.
  % \end{enumerate} 
  The category
  % 
  $\X$ is said to be \emph{$\mathcal{M}$-adhesive} if
  \begin{enumerate}
  \item it has \emph{$\mathcal{M}$-pullbacks}, i.e.~pullbacks along arrows of $\mathcal{M}$;
  \item it has \emph{$\mathcal{M}$-pushouts}, i.e.~pushouts along arrows of $\mathcal{M}$;
  \item  $\mathcal{M}$-pushouts are $\mathcal{M}$-Van Kampen squares.
  \end{enumerate}

@@

\begin{theorem}
  \label{thm:slice-functors}
  Let $\X$ be an $\mathcal{M}$-adhesive category. Given an object $X$
  the category $\X/X$ is $\mathcal{M}/X$-adhesive with
  $\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in
  \mathcal{M}\}$. Similarly, $X/\X$ is $X/\mathcal{M}$-adhesive with
  $X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in \mathcal{M}\}$.

@@

\begin{example}
  \label{ex:adhesive}
  $\cat{Set}$ is adhesive, and, more generally, every topos is
  adhesive~\cite{lack2006toposes}. By the closure properties above, every presheaf $[\cat{X},\cat{Set}]$ is adhesive, thus the category
  $\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is adhesive
  where $E \rightrightarrows {V}$ is the two objects category with two
  morphisms $s,t \colon{E} \to {V}$. Similarly, various
  categories of hypergraphs can be shown to be adhesive, such as term
  graphs and hierarchical graphs~\cite{CastelnovoGM24}. Note that the category $\cat{sGraphs}$ of simple graphs, 
  i.e.~graphs without parallel edges, is
  $\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not
  quasiadhesive.
\end{example}

@@

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair  of
  arrows $\rho = (l: K \to L, r: K \to R)$ such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

@@

  \vspace{.1cm}
  \noindent
  \parbox{10cm}{\hspace{15pt}Given two objects $G$ and $H$ and a rule
    $\rho=(l,r)$ in $\R$, a \emph{direct derivation $\mathscr{D}$ from
      $G$ to $H$ applying rule $\rho$} is a diagram as the one
    aside, in which both squares are pushouts. The arrow $m$ is called
    the \emph{match} of the derivation, while $h$ is its
    \emph{back-match}.  We denote a direct derivation $\dder{D}$
    from $G$ to $H$ as $\dder{D}\colon G\Mapsto H$. }
  \parbox{3cm}{$\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}&
      H}$}
\end{definition}

@@

%It can be seen that 
$\mathcal{M}$-adhesivity of $\X$ guarantees the
uniqueness of the result of applying a rule to an object, i.e.~two
direct derivations using the same match are abstraction equivalent
(see Proposition~\ref{prop:unique}).

@@

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  We say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  \emph{sequentially independent}
  if there is a pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes. In this case $(i_0,i_1)$ is called an  \emph{independence
    pair}.
  \[\xymatrix@R=16pt@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}
  \]
\end{definition}

@@

\begin{definition}[Switch]
  \label{def:switch}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation made
  of two sequentially independent derivations $\dder{D}_0$,
  $\dder{D}_1$ with independence pair $(i_0, i_1)$, as in the diagram
  on the left below. A \emph{switch} of $\der{D}$ along $(i_0,i_1)$ is
  a derivation $\der{E}=\{\dder{E}_i\}_{i=0}^1$, between the same objects and using the same rules
  in reverse order, as in the diagram on the right below
  \[
    \def\objectstyle{\scriptstyle}\def\labelstyle{\scriptstyle}
    \xymatrix@C=1.1mm@R=8mm{
      {L_0} \ar[d]_{m_0}
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]_{k_0}
      & & {R_0} \ar[drr]|(.35)\hole_(.6){h_0}  \ar@/^.22cm/[drrrrrr]|(.3)\hole^(.75){i_0}
      & & & & 
      % 
      {L_1}\ar[dll]|(.35)\hole^(.6){m_1} \ar@/_.22cm/_(.75){i_1}[dllllll]
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]^{k_1}
      & & {R_1} \ar[d]^{h_1} \\
      % 
      % 
      {G_0}
      & & {D_0} \ar@{>->}[ll]^{f_0} \ar[rrrr]_{g_0}
      & & & & {G_1} & &
      & &  {D_1} \ar@{>->}[llll]^{f_1} \ar[rr]_{g_1}
      & & {G_2}
    }
    % 
    \xymatrix@C=1.1mm@R=8mm{
      {L_1} \ar[d]_{m_0'}
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]_{k_0'}
      & & {R_1} \ar[drr]|(.35)\hole_(.6){h_0'}  \ar@/^.22cm/@{.>}^(.75){i_0'}|(.3)\hole[drrrrrr]
      & & & & 
      % 
      {L_0}\ar[dll]|(.35)\hole^(.6){m_1'} \ar@/_.22cm/@{.>}_(.75){i_1'}[dllllll] 
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_1'}
      & & {R_0} \ar[d]^{h_1'} \\
      % 
      % 
      {G_0}
      & & {D_0'} \ar@{>->}[ll]^{f_0'} \ar[rrrr]_{g_0'}
      & & & & {G'_1} & &
      & &  {D_1'} \ar@{>->}[llll]^{f_1'} \ar[rr]_{g_1'}
      & & {G_2}  }
    %
  \]
  such that there is an independence pair $(i_0', i_1')$ between
  $\dder{E}_0$ and $\dder{E}_1$ and 
  \begin{center}   
    $m_0=f_0' \circ i_1'$
    \qquad $h_1=g_1' \circ i_0'$
    \qquad $m_0'= f_0 \circ i_1$
    \qquad $h_1'= g_{1}\circ i_0$.
  \end{center}

@@

As we mentioned, for linear rewriting systems
%when viewing  derivations as concurrent computations, 
it is canonical to identify
derivations that are equal ``up to switching'', i.e.~that differs
only for the order of independent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

@@

    Let us denote by $\nu_{k,h}$ the composition
  $\nu_h \circ \nu_{h-1} \circ \ldots \nu_k$. We say that the
  switching sequence \emph{consists of inversions} if for all
  $k \in \interval[0]{m}$ the transposition $\nu_k$ is an inversion
  for $\nu_{1,m}$.

@@

Even if we work in strong enforcing rewriting systems where sequential
independence ensures switchability, when dealing with left-linear
rules there is a further -- and possibly more serious -- issue, 
i.e.~the fact that there can be more than one independence pair between
the same derivations. This hinders the very idea of using
sequential independence as a basis of a theory of concurrency for
rewriting systems, since exchanges performed using different
independence pairs may lead to derivations that are not abstraction
equivalent, and thus one would equate computations that should
definitively be taken apart, as shown in the example below.

\begin{example}
  Consider the derivation $\der{E}$ from Example~\ref{ex:seq-ind} (see Fig.~\ref{fi:derE}).
  %
  The last two steps are sequential independent, but one easily sees
  that there are two distinct independence pairs, as the left-hand side
  of $\rho_1$ can be mapped either to node $1$ or to node
  $2$ in $D_1'$. Correspondingly, there
  are two switches of $\der{E}$: one is the derivation $\der{D}$ in
  Fig.~\ref{fi:derD} we started from, the other is the derivation
  $\der{D}'$ in Fig.~\ref{fi:derD1}.

@@

\begin{definition}[Node-merging graphical rewriting systems]
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph
  and $\G$ the free category over $\mathcal{G}$. The
  category $\gph{G}$ of \emph{$\mathcal{G}$-graphs} is the category
  $\Set^{\G}$. A
  \emph{node-merging graphical rewriting system}\todo{altro nome? source/leaf preserving} is a left-linear
  rewriting system $(\gph{G}, \R)$ such that  the
  following hold for each rule
  $(l\colon K\to L, r\colon K\to R)$ in $\R$
  \begin{enumerate}
  \item for every $v\in V$ and $x\in L(v)$, there exists a vertex $w$
    with depth $0$ and an arrow $p\colon w\to v$ (i.e.~a path in
    $\mathcal{G}$) such that $x$ belongs to the image of
    $L(p)\colon L(w)\to L(v)$;
  \item $r\colon K\to R$ is mono on vertices of depth $0$, i.e.~for every
    $v\in V$ such that $\dph(v)=0$, the component $r_v:K(v)\to R(v)$
    is mono.
  \end{enumerate}
\end{definition}
